---
editor_options:
  markdown:
    wrap: 72
output: pdf_document
---
# Introduction
The central question I will look at is "How have NYPD shooting incidents trended over time?" I will look at how they have increased or decreased, and whether those patterns hold across NYC's five boroughs.

# Import

Begin by reading in the NYPD Shooting Data

```{r read_source}
library(tidyverse)

url = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd_shootings_raw = read_csv(url)
```

# Tidy

Now we'll do some cleanup: - Changing variable types - Getting rid of
unneeded columns

```{r tidy}
nypd_shootings_tidy <-
  nypd_shootings_raw %>% mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>% mutate(across(
    c(
      BORO,
      PRECINCT,
      JURISDICTION_CODE,
      PERP_AGE_GROUP,
      PERP_SEX,
      PERP_RACE,
      VIC_AGE_GROUP,
      VIC_SEX,
      VIC_RACE
    ),
    as.factor
  )) %>% select(
    -c(
      X_COORD_CD,
      Y_COORD_CD,
      Latitude,
      Longitude,
      Lon_Lat,
      LOC_OF_OCCUR_DESC,
      LOC_CLASSFCTN_DESC,
      LOCATION_DESC
    )
  )

summary(nypd_shootings_tidy)
```

There are some missing values, particularly in the "PERP" columns, which
likely stands for perpetrator and, understandably, may not be known in
all cases (or may be non-existent). Likely, I will remove these, but I
want to see the analysis play out a bit.

# Visualize

I'll start with a simple graph of shootings over time. To help with
this, I'll summarize the data a bit, first

```{r shootings_time_series}
nypd_shootings_tidy %>%
  mutate(OCCUR_YEAR = trunc.Date(OCCUR_DATE, "years")) %>%
  group_by(OCCUR_YEAR) %>%
  summarize(INCIDENT_COUNT = n()) %>%
  ggplot(aes(OCCUR_YEAR, INCIDENT_COUNT)) +
  labs(x = "Year", y = "Shooting Count") +
  geom_line() +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_date(date_minor_breaks = "1 years") +
  theme_minimal()
```

We'll also break this out by Borough to see how they compare to one
another:

```{r boro_time_series}
nypd_shootings_tidy %>%
  mutate(OCCUR_YEAR = trunc.Date(OCCUR_DATE, "years")) %>%
  group_by(BORO, OCCUR_YEAR) %>%
  summarize(INCIDENT_COUNT = n()) %>%
  ggplot(aes(OCCUR_YEAR, INCIDENT_COUNT, group = BORO, color = BORO)) +
  labs(x = "Year", y = "Shooting Count", color = "Borough") +
  geom_line() +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_date(date_minor_breaks = "1 years") +
  theme_minimal()
```
```{r faceted_time_series}
nypd_shootings_tidy %>% 
  mutate(occur_month = trunc.Date(OCCUR_DATE, "months")) %>% 
  group_by(occur_month, BORO) %>% 
  summarize(incident_count = n()) %>% 
  ggplot(aes(occur_month, incident_count, color = BORO)) + 
  facet_wrap(~BORO) + 
  geom_line() + 
  geom_smooth(mapping = aes(color = NULL), span = 0.25) + 
  scale_x_date(date_minor_breaks = "1 years") + 
  theme_minimal()
```
# Analysis

From the year 2000 to 2020, shooting incidents had generally decreased,
but there was a sharp uptick in the year 2020, with shootings returning
to 2000 levels. It's tempting to attribute that sudden increase to
COVID-19, but there are some interesting counterpoints that emerge when
we look at the data by borough.

Queens and Brooklyn fit the overall trend: downward trajectory in
shootings with a return to 2000 levels in 2020. But Manhattan and the
Bronx are somewhat different. They rose to higher levels than the year
2000. Furthermore, they hit their peaks a year later, in 2021.

This begs some questions: \* Is COVID-19 to blame for the increase in
shootings in 2020 and 2021? \* If so, was there some difference in
timing with respect to how COVID-19 affected Manhattan and Bronx vs.
Brooklyn and Queens that might account for the delayed increase in
shooting incidents in the latter? \* If COVID-19 *is* to blame, in what
way? Were there changes in policing policy? \* Were there other
significant events that occurred around 2020-2021 that might account for
the upticks: changes in policing laws, changes in government leadership,
etc.?

And, although I didn't look at some of the other factors, further
analyses might drill into factors that tell us more about the nature of
the crime:

-   Was there a change in the distribution of ages of victims &
    perpetrators that might suggest a shift from street crimes toward
    domestic crimes (or vice versa)?
-   What about the races of victims and perpetrators, again, suggesting
    a change in the motive or nature of the crime, or a difference in
    the way external influences like COVID or policy changes affected
    people of different races?

# Addressing Bias

Bias was foremost in my mind when I first looked at this data set. Aside
from geographical data, the factors included several that could
introduce bias, including race and sex. The initial questions that
jumped to mind were: \* Are there correlations between race and shooting
incidents? \* Are people in one racial class more likely to perpetrate
crime on another?

These questions speak to a bias that suggests that race correlates with
violence. The limitations of this data set could aggravate an analysis
around that subject. It would be important to consider the population
makeup of the various boroughs and precincts, as well as other factors
such as socioeconomic status.

Aware of the complexities, for this analysis, I attempted to mitigate
bias in my analysis by focusing on something a little less prone to
bias: municipalities and time. Admittedly, even something that *looks*
unlikely to introduce bias, such as municipality, can hide factors like
race, which can correlate to municipalitiy as people tend to congregate
with others who share a similar cultural, socioeconomic, and racial
backgrounds.

# Epilogue

Here's my setup

```{r epilogue, echo=FALSE}
sessionInfo()
```
